---
title: "Neonicotinoids in Wisconsin Waters"
subtitle: "Exploratory analysis of DATCP surface water and groundwater monitoring data"
date: today
format:
  html:
    pagetitle: "Neonicotinoids in Wisconsin Waters — Analysis"
    theme: cosmo
    css: analysis.css
    toc: true
    toc-depth: 2
    toc-expand: true
    number-sections: true
    fig-align: center
    fig-width: 9
    fig-height: 5
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

library(tidyverse)
library(sf)
library(gt)
library(mapgl)
library(plotly)

knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))

load("app/data.RData")
source("app/functions.R")

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = brand$grid, linewidth = 0.3),
      strip.text = element_text(face = "bold", size = 12, hjust = 0),
      plot.caption = element_text(color = brand$muted, hjust = 0),
      legend.position = "top",
      legend.justification = "left"
    )
)

# featured analytes + combined total (D3, D4)
res_f <- featured_results(results, samples)

# flag each benchmark as primary for its water type (D6)
exc <- exceedances |>
  left_join(distinct(benchmarks, benchmark, short_label, primary_for), join_by(benchmark)) |>
  mutate(primary = as.character(site_type) == primary_for)

bm_labels <- benchmark_labels(benchmarks)

span_years <- function(d) paste0(min(year(d)), "–", max(year(d)))

log_breaks <- c(0.001, 0.003, 0.01, 0.03, 0.1, 0.3, 1, 3, 10, 30, 100, 300, 1000)
```

::: {.callout-note}
## About this report

This report explores neonicotinoid insecticide monitoring data collected by the Wisconsin Department of Agriculture, Trade and Consumer Protection (DATCP) in surface water (streams and rivers) and groundwater (private and monitoring wells). It prototypes the analyses and figures for the public dashboard.

**Analytes.** Samples were screened for six neonicotinoids. Three are detected often and are the focus of this report: **imidacloprid, clothianidin and thiamethoxam**. **Dinotefuran** was detected only `r sum(results$detected & results$analyte == "Dinotefuran")` times, and **acetamiprid** and **thiacloprid** were screened for but never detected.

**Conventions** (details in `PLAN.md`):

- A result is a *detection* when the lab reported a concentration above zero. Non-detects are never treated as zero in displays.
- Concentration statistics (median, mean, max) are computed **over detections only**.
- The *combined total* for a sample is the sum of all detected neonicotinoid concentrations, with non-detects counted as zero. It has no benchmark.
- Surface water and groundwater are always summarized separately.
- All concentrations are in µg/L (equivalent to parts per billion, ppb).
- Private well locations are shown at approximate (≈1 km) positions.
:::

# Data overview

```{r}
overview <- samples |>
  add_site_geo(sites, "county") |>
  summarize(
    Sites = n_distinct(site_id),
    `Sample events` = n(),
    Counties = n_distinct(county),
    Years = span_years(date),
    .by = site_type
  )

overview |>
  gt(rowname_col = "site_type") |>
  fmt_number(c(Sites, `Sample events`, Counties), decimals = 0) |>
  tab_options(table.width = pct(70))
```

A *sample event* is one site sampled on one date (one well was sampled twice on the same day; both samples are kept). Each event was analyzed for up to six neonicotinoids.

## Analyte screen

```{r}
results |>
  summarize_detections(site_type, analyte) |>
  select(site_type, analyte, n_samples, n_detected, det_freq, max_det) |>
  arrange(site_type, desc(n_detected)) |>
  gt(groupname_col = "site_type", rowname_col = "analyte") |>
  cols_label(
    n_samples = "Samples analyzed",
    n_detected = "Detections",
    det_freq = "Detection frequency",
    max_det = "Max detected (µg/L)"
  ) |>
  fmt_number(c(n_samples, n_detected), decimals = 0) |>
  fmt_percent(det_freq, decimals = 1) |>
  fmt(max_det, fns = \(x) fmt_conc(x, units = FALSE)) |>
  sub_missing(missing_text = "–") |>
  tab_options(table.width = pct(80))
```

## Sampling effort over time

```{r}
#| fig-height: 4
samples |>
  count(site_type, year) |>
  ggplot(aes(year, n)) +
  geom_col(fill = brand$red, width = 0.8) +
  facet_wrap(~site_type, scales = "free_y") +
  labs(x = NULL, y = "Sample events")
```

Groundwater sampling before 2011 is sparse (`r nrow(filter(samples, site_type == "Groundwater", year < 2011))` events). The monitoring network grew substantially from 2015 onward.

## Seasonality (surface water)

```{r}
#| fig-height: 4
res_f |>
  filter(site_type == "Surface water", analyte == "Total") |>
  summarize_detections(month) |>
  ggplot(aes(month, det_freq)) +
  geom_col(fill = analyte_colors[["Total"]], width = 0.8) +
  geom_text(aes(label = paste0("n=", n_samples)), vjust = -0.4, size = 3, color = brand$muted) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = scales::percent, expand = expansion(c(0, 0.1))) +
  labs(x = NULL, y = "Samples with any detection", caption = "Surface water, any neonicotinoid detected, by sampling month.")
```

# Detection limits

Laboratory detection limits (DL) have dropped roughly 20–50-fold over the monitoring period. This matters for two reasons:

1. **Trends.** A higher share of samples shows detections in recent years partly because the lab can now see lower concentrations (see @sec-trends).
2. **Benchmarks.** A non-detect with a DL above a benchmark can't be classified as above or below it (see @sec-exceed).

```{r}
#| fig-height: 4.5
results |>
  filter(featured) |>
  count(site_type, year, dl) |>
  ggplot(aes(year, factor(dl), fill = n)) +
  geom_tile(color = "white", linewidth = 0.6) +
  facet_wrap(~site_type) +
  scale_fill_gradientn(colors = freq_ramp, trans = "log10", labels = scales::comma) +
  labs(x = NULL, y = "Detection limit (µg/L)", fill = "Results", caption = "Imidacloprid, clothianidin and thiamethoxam results by year and detection limit.") +
  theme(legend.key.width = unit(2, "cm"))
```

# Detection frequency

```{r}
det_by_analyte <- res_f |>
  summarize_detections(site_type, analyte)
```

```{r}
#| fig-height: 4.5
det_by_analyte |>
  ggplot(aes(analyte, det_freq, fill = analyte)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = fmt_pct(det_freq, 0)), vjust = -0.4, size = 3.5) +
  facet_wrap(~site_type) +
  scale_fill_manual(values = analyte_colors, guide = "none") +
  scale_y_continuous(labels = scales::percent, expand = expansion(c(0, 0.1))) +
  labs(x = NULL, y = "Samples with a detection", caption = "'Total' = any of the six neonicotinoids detected in the sample.")
```

```{r}
det_by_analyte |>
  mutate(pct_sites = n_sites_detected / n_sites) |>
  select(site_type, analyte, n_samples, det_freq, n_sites, n_sites_detected, pct_sites) |>
  gt(groupname_col = "site_type", rowname_col = "analyte") |>
  cols_label(
    n_samples = "Samples",
    det_freq = "% samples detected",
    n_sites = "Sites",
    n_sites_detected = "Sites with ≥1 detection",
    pct_sites = "% sites"
  ) |>
  fmt_number(c(n_samples, n_sites, n_sites_detected), decimals = 0) |>
  fmt_percent(c(det_freq, pct_sites), decimals = 1) |>
  tab_options(table.width = pct(85))
```

## Trends over time {#sec-trends}

Detection frequency by year, shown three ways:

- **As reported:** each result compared to the detection limit in use at the time.
- **Common reporting levels:** results re-censored at 0.067 µg/L and at 0.2 µg/L. Samples analyzed with a higher DL are dropped and detections below the level count as non-detects.

The common-level lines compare like with like across years, at the cost of shorter records.

```{r}
#| fig-height: 8
trend <- bind_rows(
  res_f |> filter(analyte != "Total") |> mutate(basis = "As reported"),
  res_f |> filter(analyte != "Total") |> apply_reporting_level(0.067) |> mutate(basis = "Common level ≥ 0.067 µg/L"),
  res_f |> filter(analyte != "Total") |> apply_reporting_level(0.2) |> mutate(basis = "Common level ≥ 0.2 µg/L")
) |>
  mutate(basis = fct_inorder(basis)) |>
  summarize_detections(site_type, analyte, year, basis) |>
  filter(n_samples >= 20, year >= 2011)

trend |>
  ggplot(aes(year, det_freq, color = analyte)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_grid(basis ~ site_type) +
  scale_color_manual(values = analyte_colors) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(2011, 2025, 2)) +
  labs(x = NULL, y = "Samples with a detection", color = NULL, caption = "Years with fewer than 20 samples omitted.") +
  theme(strip.text.y = element_text(angle = 0, hjust = 0, size = 10))
```

# Concentrations

Concentrations of **detections only**, on a log scale. Dashed lines show the benchmarks that apply to each water type: EPA aquatic life benchmarks for surface water, and the proposed Wisconsin NR 140 groundwater standards for groundwater.

```{r}
#| fig-height: 5.5
det_pts <- res_f |>
  filter(detected, analyte != "Total")

bm_lines <- benchmarks |>
  filter(analyte %in% featured_analytes) |>
  mutate(site_type = factor(primary_for, site_types), analyte = factor(analyte, names(analyte_colors))) |>
  filter(value <= 20)

off_scale <- benchmarks |>
  filter(analyte %in% featured_analytes, value > 20) |>
  mutate(txt = paste0(analyte, " ", short_label, " (", fmt_conc(value), ")")) |>
  pull(txt)

det_pts |>
  ggplot(aes(analyte, result)) +
  geom_jitter(aes(color = analyte), width = 0.25, height = 0, alpha = 0.35, size = 1.2) +
  geom_boxplot(fill = NA, outlier.shape = NA, width = 0.5, linewidth = 0.4) +
  geom_segment(
    data = bm_lines,
    aes(x = as.numeric(analyte) - 0.45, xend = as.numeric(analyte) + 0.45, y = value, yend = value),
    linetype = "dashed", linewidth = 0.5, color = brand$text
  ) +
  geom_label(
    data = bm_lines,
    aes(x = as.numeric(analyte) - 0.45, y = value, label = short_label),
    hjust = 0, vjust = -0.2, size = 2.7, color = brand$text,
    fill = alpha("white", 0.85), label.size = 0, label.padding = unit(0.1, "lines")
  ) +
  facet_wrap(~site_type) +
  scale_color_manual(values = analyte_colors, guide = "none") +
  scale_y_log10(breaks = log_breaks, labels = \(x) format(x, scientific = FALSE, drop0trailing = TRUE)) +
  labs(
    x = NULL, y = "Detected concentration (µg/L)",
    caption = str_wrap(paste("Off scale (above 20 µg/L):", paste(off_scale, collapse = "; ")), 130)
  )
```

```{r}
det_by_analyte |>
  select(site_type, analyte, n_detected, median_det, mean_det, max_det) |>
  gt(groupname_col = "site_type", rowname_col = "analyte") |>
  cols_label(
    n_detected = "Detections",
    median_det = "Median",
    mean_det = "Mean",
    max_det = "Max"
  ) |>
  tab_spanner("Concentration of detections (µg/L)", c(median_det, mean_det, max_det)) |>
  fmt_number(n_detected, decimals = 0) |>
  fmt(c(median_det, mean_det, max_det), fns = \(x) fmt_conc(x, units = FALSE)) |>
  tab_options(table.width = pct(70))
```

## Highest concentrations

```{r}
res_f |>
  filter(detected, analyte != "Total") |>
  slice_max(result, n = 1, by = c(site_type, site_id), with_ties = FALSE) |>
  slice_max(result, n = 8, by = site_type) |>
  add_site_geo(sites, c("site_label", "county")) |>
  transmute(site_type, site_label, county, date, analyte, result) |>
  gt(groupname_col = "site_type") |>
  cols_label(site_label = "Site", county = "County", date = "Date", analyte = "Analyte", result = "Result (µg/L)") |>
  fmt(result, fns = \(x) fmt_conc(x, units = FALSE)) |>
  tab_options(table.width = pct(90))
```

The highest single result at each site, top 8 sites per water type.

# Benchmark exceedances {#sec-exceed}

Each result is compared to every benchmark available for its analyte:

| Benchmark | Source | Primary comparison for |
|---|---|---|
| Aquatic acute / chronic (freshwater invertebrates) | [US EPA OPP Aquatic Life Benchmarks](https://www.epa.gov/pesticide-science-and-assessing-pesticide-risks/aquatic-life-benchmarks-and-ecological-risk) | Surface water |
| Proposed enforcement standard (ES) / preventive action limit (PAL) | [Wisconsin DNR NR 140, Cycle 13 review (proposed)](https://dnr.wisconsin.gov/topic/Groundwater/NR140.html) | Groundwater |

```{r}
benchmarks |>
  select(analyte, short_label, value) |>
  pivot_wider(names_from = short_label, values_from = value) |>
  gt(rowname_col = "analyte") |>
  fmt(-analyte, fns = \(x) fmt_conc(x, units = FALSE)) |>
  sub_missing(missing_text = "–") |>
  tab_header(title = "Benchmark values (µg/L)") |>
  tab_options(table.width = pct(70))
```

Each result is classified as:

- **Exceeds:** detected at or above the benchmark.
- **Below:** detected below the benchmark, or not detected with a detection limit at or below the benchmark.
- **Indeterminate:** not detected, but the detection limit was **above** the benchmark, so we can't tell. This happens mostly with older samples and the very low imidacloprid and clothianidin chronic benchmarks.

::: {.callout-important}
Exceeding an aquatic life benchmark means the concentration is at a level that may harm sensitive aquatic invertebrates *if it persists in surface water*. Applied to groundwater, it indicates potential risk where groundwater discharges to streams, not a drinking-water risk. The NR 140 values are *proposed* groundwater quality standards, and are not yet enforceable.
:::

## Summary table

```{r}
exc_summary <- exc |>
  filter(analyte %in% featured_analytes) |>
  summarize_exceedances(site_type, analyte, benchmark, short_label, value, primary) |>
  arrange(site_type, desc(primary), analyte, benchmark)

exc_summary |>
  mutate(
    group = paste0(site_type, if_else(primary, " — primary benchmarks", " — other benchmarks")),
    exceed_txt = sprintf("%s (%s)", fmt_n(n_exceeds), fmt_pct(pct_exceeds)),
    indet_txt = sprintf("%s (%s)", fmt_n(n_indeterminate), fmt_pct(pct_indeterminate)),
    sites_txt = sprintf("%s of %s", fmt_n(n_sites_exceed), fmt_n(n_sites))
  ) |>
  select(group, analyte, short_label, value, n_samples, exceed_txt, indet_txt, sites_txt, n_exceeds) |>
  gt(groupname_col = "group") |>
  cols_hide(n_exceeds) |>
  cols_label(
    analyte = "Analyte",
    short_label = "Benchmark",
    value = "µg/L",
    n_samples = "Samples",
    exceed_txt = "Exceeds",
    indet_txt = "Indeterminate",
    sites_txt = "Sites with ≥1 exceedance"
  ) |>
  fmt(value, fns = \(x) fmt_conc(x, units = FALSE)) |>
  fmt_number(n_samples, decimals = 0) |>
  tab_style(cell_text(weight = "bold"), cells_body(exceed_txt, rows = n_exceeds > 0)) |>
  tab_options(table.width = pct(100), row_group.font.weight = "bold")
```

## Share of samples by status

```{r}
#| fig-height: 8
exc_summary |>
  arrange(site_type, analyte, desc(primary), benchmark) |>
  select(site_type, analyte, short_label, primary, n_exceeds, n_below, n_indeterminate) |>
  mutate(bm = paste0(analyte, " — ", short_label, if_else(primary, " (primary)", ""))) |>
  mutate(bm = fct_rev(fct_inorder(bm))) |>
  pivot_longer(c(n_exceeds, n_below, n_indeterminate), names_to = "status", values_to = "n") |>
  mutate(
    status = factor(status, c("n_exceeds", "n_below", "n_indeterminate"), names(status_colors)),
    status = fct_rev(status)
  ) |>
  ggplot(aes(n, bm, fill = status)) +
  geom_col(position = "fill", width = 0.7, color = "white", linewidth = 0.5) +
  facet_wrap(~site_type, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = status_colors, breaks = names(status_colors)) +
  scale_x_continuous(labels = scales::percent, expand = expansion(c(0, 0.02))) +
  labs(x = "Share of samples", y = NULL, fill = NULL)
```

## Exceedances over time

Annual share of samples in each status for the benchmarks that are exceeded most often. The shrinking *Indeterminate* share shows the effect of falling detection limits.

```{r}
#| fig-height: 8
key_combos <- exc_summary |>
  filter(n_exceeds >= 50) |>
  distinct(site_type, analyte, benchmark, short_label)

exc |>
  semi_join(key_combos, join_by(site_type, analyte, benchmark)) |>
  filter(year >= 2011) |>
  count(site_type, analyte, benchmark, short_label, year, status) |>
  arrange(site_type, analyte, benchmark) |>
  mutate(
    panel = fct_inorder(paste0(site_type, ": ", analyte, " · ", short_label)),
    status = fct_rev(status)
  ) |>
  ggplot(aes(year, n, fill = status)) +
  geom_col(position = "fill", width = 0.85) +
  facet_wrap(~panel, ncol = 2) +
  scale_fill_manual(values = status_colors, breaks = names(status_colors)) +
  scale_y_continuous(labels = scales::percent, breaks = c(0, 0.5, 1)) +
  scale_x_continuous(breaks = seq(2011, 2025, 2)) +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme(strip.text = element_text(size = 10))
```

# Groundwater context

## By well type

```{r}
gw_res <- res_f |>
  filter(site_type == "Groundwater") |>
  add_site_geo(sites, "well_use")

gw_det <- gw_res |>
  filter(analyte == "Total") |>
  summarize_detections(well_use) |>
  select(well_use, n_sites, n_samples, det_freq)

gw_exc <- exc |>
  filter(site_type == "Groundwater", primary) |>
  add_site_geo(sites, "well_use") |>
  summarize(
    n_exceed_es = sum(status == "Exceeds" & benchmark == "proposed_es"),
    n_exceed_pal = sum(status == "Exceeds" & benchmark == "proposed_pal"),
    sites_exceed_es = n_distinct(site_id[status == "Exceeds" & benchmark == "proposed_es"]),
    sites_exceed_pal = n_distinct(site_id[status == "Exceeds" & benchmark == "proposed_pal"]),
    .by = well_use
  )

gw_det |>
  left_join(gw_exc, join_by(well_use)) |>
  arrange(desc(n_sites)) |>
  gt(rowname_col = "well_use") |>
  cols_label(
    n_sites = "Wells", n_samples = "Samples", det_freq = "% samples with any detection",
    n_exceed_pal = "Samples > PAL", sites_exceed_pal = "Wells > PAL",
    n_exceed_es = "Samples > ES", sites_exceed_es = "Wells > ES"
  ) |>
  cols_move(c(n_exceed_pal, sites_exceed_pal), after = det_freq) |>
  fmt_number(-c(well_use, det_freq), decimals = 0) |>
  fmt_percent(det_freq, decimals = 1) |>
  tab_footnote("Proposed NR 140 standards; only imidacloprid results exceed them.") |>
  tab_options(table.width = pct(100))
```

## Well depth

```{r}
#| fig-height: 4
gw_res |>
  filter(analyte != "Total") |>
  add_site_geo(sites, "well_depth") |>
  filter(!is.na(well_depth)) |>
  mutate(depth_bin = cut(well_depth, c(0, 25, 50, 100, 200, Inf), labels = c("< 25", "25–50", "50–100", "100–200", "200+"))) |>
  summarize_detections(depth_bin, analyte) |>
  ggplot(aes(depth_bin, det_freq, fill = analyte)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75, color = "white", linewidth = 0.5) +
  scale_fill_manual(values = analyte_colors) +
  scale_y_continuous(labels = scales::percent, expand = expansion(c(0, 0.05))) +
  labs(x = "Well depth (ft)", y = "Samples with a detection", fill = NULL, caption = sprintf("Only wells with a recorded depth (%s of %s wells).", fmt_n(sum(!is.na(sites$well_depth))), fmt_n(sum(sites$site_type == "Groundwater"))))
```

# Geography

## County and watershed summaries

```{r}
geo_summary <- function(geo_col) {
  det <- res_f |>
    filter(analyte == "Total") |>
    add_site_geo(sites, geo_col) |>
    summarize_detections(site_type, .data[[geo_col]]) |>
    select(site_type, all_of(geo_col), n_sites, n_samples, det_freq, median_det)

  ex <- exc |>
    filter(primary) |>
    add_site_geo(sites, geo_col) |>
    summarize(
      n_sites_exceed = n_distinct(site_id[status == "Exceeds"]),
      .by = c(site_type, all_of(geo_col))
    )

  left_join(det, ex, join_by(site_type, !!geo_col)) |>
    mutate(pct_sites_exceed = n_sites_exceed / n_sites)
}

county_summary <- geo_summary("county")
wshed_summary <- geo_summary("wshed_code")
```

`r n_distinct(county_summary$county[county_summary$site_type == "Groundwater"])` of 72 counties have groundwater data and `r n_distinct(county_summary$county[county_summary$site_type == "Surface water"])` have surface water data. `r n_distinct(wshed_summary$wshed_code[wshed_summary$site_type == "Groundwater"])` of 334 DNR watersheds have groundwater data and `r n_distinct(wshed_summary$wshed_code[wshed_summary$site_type == "Surface water"])` have surface water data.

```{r}
choropleth <- function(shapes, summary, key, value, ramp, label, pct = TRUE) {
  crossing(shapes, site_type = factor(site_types, site_types)) |>
    st_as_sf() |>
    left_join(summary, join_by(!!key, site_type)) |>
    ggplot() +
    geom_sf(aes(fill = .data[[value]]), color = "#c3c2b7", linewidth = 0.15) +
    geom_sf(data = wi_state, fill = NA, color = brand$text, linewidth = 0.3) +
    facet_wrap(~site_type) +
    scale_fill_gradientn(
      colors = ramp,
      na.value = "white",
      labels = if (pct) scales::percent else waiver(),
      limits = if (pct) c(0, 1) else NULL
    ) +
    labs(fill = label, caption = "White = no monitoring data.") +
    theme_void() +
    theme(
      legend.position = "top",
      legend.key.width = unit(2, "cm"),
      strip.text = element_text(face = "bold", size = 12),
      plot.caption = element_text(color = brand$muted)
    )
}
```

### Detection frequency by county

```{r}
#| fig-height: 5.5
choropleth(counties, county_summary, "county", "det_freq", freq_ramp, "Samples with any neonicotinoid detected")
```

### Detection frequency by DNR watershed

```{r}
#| fig-height: 5.5
choropleth(watersheds, wshed_summary, "wshed_code", "det_freq", freq_ramp, "Samples with any neonicotinoid detected")
```

### Sites exceeding a primary benchmark, by county

```{r}
#| fig-height: 5.5
choropleth(counties, county_summary, "county", "pct_sites_exceed", conc_ramp, "Sites with ≥1 primary benchmark exceedance")
```

Surface water sites are compared to the EPA aquatic benchmarks and groundwater sites to the proposed NR 140 standards. Almost all of these exceedances are imidacloprid or clothianidin chronic aquatic benchmarks (surface water), or imidacloprid PAL/ES (groundwater).

```{r}
county_summary |>
  filter(n_samples >= 30) |>
  slice_max(det_freq, n = 10, by = site_type) |>
  select(site_type, county, n_sites, n_samples, det_freq, n_sites_exceed) |>
  gt(groupname_col = "site_type", rowname_col = "county") |>
  cols_label(n_sites = "Sites", n_samples = "Samples", det_freq = "% samples detected", n_sites_exceed = "Sites exceeding a primary benchmark") |>
  fmt_percent(det_freq, decimals = 1) |>
  tab_header(title = "Counties with the highest detection frequency (≥ 30 samples)") |>
  tab_options(table.width = pct(80))
```

## Interactive site map

Sites colored by their most serious result: exceeded a primary benchmark, detected, or never detected. Click a site for its full analyte screen. Private wells are shown at approximate locations.

```{r}
#| fig-height: 7
site_status <- exc |>
  filter(primary) |>
  summarize(any_exceed = any(status == "Exceeds"), .by = c(site_type, site_id))

map_pts <- sites |>
  left_join(site_status, join_by(site_type, site_id)) |>
  left_join(build_site_popups(sites, site_analytes), join_by(site_type, site_id)) |>
  mutate(
    status = case_when(
      coalesce(any_exceed, FALSE) ~ "Exceeded a primary benchmark",
      n_detected > 0 ~ "Detected",
      .default = "Not detected"
    ),
    tooltip = paste0("<b>", site_label, "</b><br>", status)
  ) |>
  arrange(desc(status == "Not detected")) |> # draw detections on top
  st_as_sf(coords = c("map_lon", "map_lat"), crs = 4326)

status_vals <- c("Exceeded a primary benchmark", "Detected", "Not detected")
status_cols <- c(status_colors[["Exceeds"]], analyte_colors[["Imidacloprid"]], "#b9b8b1")

add_sites <- function(map, type, id) {
  map |>
    add_circle_layer(
      id = id,
      source = filter(map_pts, site_type == type),
      circle_color = match_expr("status", values = status_vals, stops = status_cols, default = "#b9b8b1"),
      circle_radius = if (type == "Surface water") 6 else 4,
      circle_stroke_color = "white",
      circle_stroke_width = 1,
      circle_opacity = 0.9,
      tooltip = "tooltip",
      popup = "popup"
    )
}

maplibre(style = carto_style("positron"), bounds = wi_state) |>
  add_line_layer(id = "Counties", source = counties, line_color = "#9a9a9a", line_width = 0.5) |>
  add_line_layer(id = "DNR watersheds", source = watersheds, line_color = "#6da7ec", line_width = 0.5, visibility = "none") |>
  add_sites("Groundwater", "Groundwater") |>
  add_sites("Surface water", "Surface water") |>
  add_categorical_legend(legend_title = "Site status", values = status_vals, colors = status_cols, circular_patches = TRUE) |>
  add_layers_control(position = "top-left", layers = c("Surface water", "Groundwater", "Counties", "DNR watersheds")) |>
  add_navigation_control() |>
  add_fullscreen_control()
```

# Combined total

The combined total adds together all detected neonicotinoids in a sample. There is no benchmark for the total, so it is shown for context only.

```{r}
#| fig-height: 4
samples |>
  filter(any_detected) |>
  mutate(n_detected = factor(pmin(n_detected, 3), 1:3, c("1 analyte", "2 analytes", "3+ analytes"))) |>
  ggplot(aes(total_conc, fill = n_detected)) +
  geom_histogram(bins = 40, color = "white", linewidth = 0.2) +
  facet_wrap(~site_type, scales = "free_y") +
  scale_x_log10(breaks = log_breaks, labels = \(x) format(x, scientific = FALSE, drop0trailing = TRUE)) +
  scale_fill_manual(values = c("#9ec5f4", "#3987e5", "#104281")) +
  labs(x = "Combined total of detections (µg/L)", y = "Samples", fill = "Detected in sample:")
```

```{r}
samples |>
  filter(any_detected) |>
  count(site_type, analytes_detected) |>
  mutate(pct = n / sum(n), .by = site_type) |>
  slice_max(n, n = 6, by = site_type) |>
  gt(groupname_col = "site_type", rowname_col = "analytes_detected") |>
  cols_label(n = "Samples", pct = "% of samples with detections") |>
  fmt_percent(pct, decimals = 1) |>
  tab_header(title = "Most common combinations of detected analytes") |>
  tab_options(table.width = pct(70))
```

# Site profile prototype

This prototypes the app's site time series. Filled points are detections. Open points are non-detects, plotted at their detection limit. Dashed lines are the primary benchmarks for the water type. Hover for details.

```{r}
site_timeseries <- function(type, id) {
  site <- filter(sites, site_type == type, site_id == id)
  d <- res_f |>
    filter(site_type == type, site_id == id, analyte != "Total") |>
    mutate(
      y = coalesce(result, dl),
      shape = if_else(detected, "Detected", "Not detected (at DL)"),
      text = sprintf("%s<br>%s<br>%s", format(date, "%b %d, %Y"), analyte, if_else(detected, fmt_conc(result), paste("ND (<", fmt_conc(dl), ")")))
    )
  bm <- benchmarks |>
    filter(analyte %in% unique(d$analyte[d$detected]), primary_for == type) |>
    mutate(analyte = factor(analyte, names(analyte_colors)))

  p <- ggplot(d, aes(date, y, color = analyte, text = text)) +
    geom_hline(data = bm, aes(yintercept = value, color = analyte), linetype = "dashed", linewidth = 0.4) +
    geom_point(aes(shape = shape), size = 2.2, stroke = 0.8) +
    scale_shape_manual(values = c("Detected" = 16, "Not detected (at DL)" = 1)) +
    scale_color_manual(values = analyte_colors) +
    scale_y_log10() +
    labs(x = NULL, y = "Concentration (µg/L)", color = NULL, shape = NULL, title = site$site_label)

  ggplotly(p, tooltip = "text") |>
    layout(legend = list(orientation = "h", y = -0.15)) |>
    config(displayModeBar = FALSE)
}

top_site <- function(type) {
  samples |>
    filter(site_type == type) |>
    summarize(n = sum(any_detected), .by = site_id) |>
    slice_max(n, n = 1, with_ties = FALSE) |>
    pull(site_id)
}
```

### Surface water example

```{r}
site_timeseries("Surface water", top_site("Surface water"))
```

### Groundwater example

```{r}
site_timeseries("Groundwater", top_site("Groundwater"))
```

# Headline numbers

Candidate values for the dashboard's introduction page stat cards.

```{r}
headline <- samples |>
  summarize(
    `Sample events` = fmt_n(n()),
    Sites = fmt_n(n_distinct(site_id)),
    Years = span_years(date),
    `Samples with any neonicotinoid` = fmt_pct(mean(any_detected), 0),
    .by = site_type
  ) |>
  left_join(
    det_by_analyte |>
      filter(analyte != "Total") |>
      slice_max(det_freq, n = 1, by = site_type) |>
      transmute(site_type, `Most detected` = sprintf("%s (%s)", analyte, fmt_pct(det_freq, 0))),
    join_by(site_type)
  ) |>
  left_join(
    exc |>
      filter(primary) |>
      summarize(
        exceed = any(status == "Exceeds"),
        .by = c(site_type, site_id, date, sample_seq)
      ) |>
      summarize(`Samples exceeding a primary benchmark` = fmt_pct(mean(exceed), 0), .by = site_type),
    join_by(site_type)
  )

headline |>
  pivot_longer(-site_type) |>
  pivot_wider(names_from = site_type) |>
  gt(rowname_col = "name") |>
  tab_options(table.width = pct(70))
```

# Appendix: data validation

```{r}
tibble(
  Check = c(
    "Exact duplicate rows removed",
    "Same-day repeat samples kept",
    "Results reported below the detection limit (counted as estimated detections)",
    "Sites outside WI boundary polygons (assigned to nearest county/watershed)",
    "Private wells with approximate locations",
    "…moved to stay inside their true county and watershed",
    "Median / max display offset of private wells (m)"
  ),
  Value = c(
    validation$exact_duplicates_removed,
    nrow(validation$repeat_same_day_samples),
    nrow(validation$estimated_below_dl),
    nrow(validation$sites_outside_wi),
    validation$blur$n_private_wells,
    validation$blur$n_moved_to_zone,
    sprintf("%.0f / %.0f", validation$blur$offset_m[["Median"]], validation$blur$offset_m[["Max."]])
  ) |> as.character()
) |>
  gt() |>
  tab_options(table.width = pct(80))
```
